Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-02-22 ms.tex
blah

Even if SN Ia are rare near the stagnation radius, they may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between subsequent supernovae, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according
to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm
  Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle
capable of dynamically clearing out gas from radii $\sim r_{\rm Ia}
\gtrsim r_{\rm s}$, at least for low mass BHs with $\sigma \ll v^{\rm
  Ia}_{w}$.  Thus, even in cases when heating is sufficiently weak
that the stagnation radius formally exceeds $r_{\rm Ia}$, the
accretion rate is limited to a maximum value,
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}g}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.  Also note that Ia supernovae only cap the accretion rate to this maximum value if the gas feeding the black hole is supplied by local stellar mass loss: the gaseous mass can become much greater if the nucleus is fed externally, e.g. due to a galaxy merger or to cooling flows developing within the galaxy over much longer timescales (e.g.~\citealt{Ciotti&Ostriker07}).

@@

\subsection{Combined Heating Rate} 
\label{sec:combined}

The total gas heating rate, 
\be
\vwO = \sqrt{(v_{w}^{\star})^{2} + (v_{w}^{\rm MSP})^{2} + (v_{w}^{\rm Ia})^{2} + (v_{w}^{\bullet})^{2}},
\label{eq:vtot}
\ee
includes contributions from stellar winds, supernovae, pulsars, and radiative SMBH feedback (we neglect kinetic SMBH feedback for the moment, given its uncertain efficacy).  The strength of each heating source depends explicitly on the SMBH mass and the stellar population in the galactic nuclear region.  The latter could best be described by a single star burst formation in the past, or by a more continuous star formation history that itself varies systematically with the galaxy mass and hence $\Mbh$.

Figure~\ref{fig:vwSourcesImp} shows the contributions of various heating sources as a function of time $\tau_{\star}$ after a burst of star formation for black holes of mass $\Mbh = 10^{6} \Msun$ (top panel) and $\Mbh = 10^{8} \Msun$ (bottom panel). The heating and mass input due to stellar winds is calculated as described in Appendix~\ref{app:windheat} (Fig.~\ref{fig:vwImp}).  The heating from SN Ia ($v_{w}^{\rm Ia}$) and Compton heating ($v_{\rm w}^{\bullet}$) are calculated from equations~\eqref{eq:vIa} and~\eqref{eq:vC} respectively using the mass input parameter from stellar winds, $\eta$ (Fig.~\ref{fig:vwImp}).  SN Ia heating also depends on a convolution of the delay time distribution (DTD; eq. [\ref{eq:DTD}]) and the star formation history, which reduces to the DTD itself for a single star burst.  We crudely account for the expected reduction in Ia heating efficiency due to the non-steady nature of the SN Ia by attenuating the Ia heating rate by a factor $e^{-\rs/\rIa}$, that reduces to zero when the stagnation radius exceeds the radius, $r_{\rm Ia}$, at which the time between subsequent supernovae is equal to the local dynamical time (eq.~[\ref{eq:rIa}]).  Finally, note that $v_{\rm w}^{\bullet}$ and $v_{\rm w}^{\rm Ia}$ also depend on the total heating rate (eq.~[\ref{eq:vtot}]), requiring us to solve an implicit equation.  Heating from MSPs is found to be negligible compared to other sources of heating for fiducial parameters and hence is not plotted for clarity.

Figure~\ref{fig:vwSourcesImp} shows that for low mass black holes, stellar heating is the most important on timescales less than a few tens of millions of years.  Compton heating comes to dominate at later times due to the high stellar mass loss rates coupled with the higher accretion rates that accompany an overall decrease in the heating rate.  For higher mass black holes, Compton heating dominates at all epochs.  Because the total heating rate declines with time, after some point the flow becomes thermally unstable near the stagnation radius, and hence across all radii.  The epoch after this occurs according to the criterion $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}| \lesssim 1$ (eq.~[\ref{eq:cooling3}]) are shown as dashed lines in Figure~\ref{fig:vwSourcesImp}.  This transition is calculated not include Compton heating, since the latter cannot truly stabilize flow due to its inability to respond instantaneously to local changes in the thermal properties of the gas.  We thus see that thermally unstable flow is the norm in the single starburst case, except at very early times ($\lsim 10^7$ years) for small $\Mbh\simeq 10^6 \Msun$.

@@

  \begin{enumerate}
  %Younger population=>higher accretion rate; generally higher eta and
  %lower vw except for very young stellar populations.
  \item We find that $\eddr \propto \Mbh^{\alpha}$, where the value of
    $\alpha$ depends on the slope of the stellar density profile
    ($\alpha\simeq0.5,1$ for cusp and core galaxies respectively). One
    possible solution that the wind mass loss rate (i.e. $\eta$)
    increases with $\Mbh$, as would be expected if low mass SMBH
    preferentially hosted younger stellar populations. Another possible
    explanation is that the effective heating rate of the CNM $v_{w}$
    decreases with decreasing SMBH mass. This could again be due to an
    older stellar population, since $v_{w} \propto \eta^{−1/2}$
    for heating sources such as SN Ia and millisecond pulsars.
  \item We compare the $\Mdot$ for each of our solution to the one
    calculated from the Bondi formula. We find that they generally to
    a factor of a few. In our model the accretion rate is determined
    from the gas properties at the stagnation radius, $\rs$, which
    agrees with the Bondi radius, $\rb$, within a factor of a
    few. Thus, use of the Bondi formula is a good approximation.
  \item We compare our solutions for density and temperature to those
    inferred from Chandra x-ray observations by \citet{AllenDunn+:2006a}
    and \citet{RussellMcNamara+:2013a} for a handful of galaxies in
    which our sample overlap with theirs. In general, we cannot match
    the observed gas density slope on large scales. 
  \end{enumerate}

>>>>

Even if SN Ia are rare near the stagnation radius, they may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between subsequent supernovae, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according
to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm
  Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle
capable of dynamically clearing out gas from radii $\sim r_{\rm Ia}
\gtrsim r_{\rm s}$, at least for low mass BHs with $\sigma \ll v^{\rm
  Ia}_{w}$.  Thus, even in cases when heating is sufficiently weak
that the stagnation radius formally exceeds $r_{\rm Ia}$, the
accretion rate is limited to a maximum value,
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.  Also note that Ia supernovae only cap the accretion rate to this maximum value if the gas feeding the black hole is supplied by local stellar mass loss: the gaseous mass can become much greater if the nucleus is fed externally, e.g. due to a galaxy merger or to cooling flows developing within the galaxy over much longer timescales (e.g.~\citealt{Ciotti&Ostriker07}).

@@

\subsection{Combined Heating Rate for Representative Star Formation History} 
\label{sec:combined}

The total gas heating rate, $\vwO$, includes contributions from
stellar winds, supernovae, pulsars, and SMBH feedback.  The strength
of each heating source will depend explicitely on the SMBH mass and
the star formation hstory in the galactic nuclear region, the latter
of which will also vary systematically with the galaxy mass and hence
with $\Mbh$.

Figure~\ref{fig:vwSourcesImp} shows the contributions of various
heating sources as a function of time $\tau_{\star}$ after a burst of
star formation for a black hole masses of $\Mbh = 10^{6} \Msun$ (top
panel) and $\Mbh = 10^{8} \Msun$ (bottom panel). The stellar wind
heating is calculated as described in appendix~\ref{app:windheat}. The
and Ia heating ($v_{w}^{\rm Ia}$) and compton heating ($v_{\rm
  w}^{\bullet}$) are calculated using equations~\eqref{eq:vIa}
and~\eqref{eq:vC} respectively. Both heating sources depend on the
value of the mass return parameter, $\eta$. We have plotted this for a
burst of star formation in Figure~\ref{fig:vwImp}.  The Ia heating rate
also depends be a convolution of the delay time distribution
(eq. [\ref{eq:DTD}]) and the star formation history. This convolution
reduces to the delay time distribution itself for a burst of star
formation. To account for the non-steady nature of the Ia heating we
multiply the Ia heating by $e^{-\rs/\rIa}$, where $\rIa$ is the radius
where the time between subsequent supernovae is equal to the local
dynamical time. The heating contribiution from
Ia supernovae will be suppressed if the time between successive
supernovae is much greater than the dynamical time at $\rs$. Note that
both $v_{\rm w}^{\bullet}$ and $v_{\rm w}^{\rm Ia}$ depend on the
total heating rate (all of the heating sources added in quadrature),
and so we have to solve a system of impllicit eqautions for these.
The MSP contribution is negligible for both cases and is not plotted.

The overall heating rate declines with time and after some time the
flow would become thermally unstable. The epoch when this occurs is
shown by dashed lines in Figure~\ref{fig:vwSourcesImp}. In calculating
this transition we have not included the compton heating. However,
this would not be able stabilize flow as it would not respond
immediately to cooling instabilities. Thus, if Compton
heating is the dominant source of heating we do not expect a steady,
thermally stable flow. Thermal instability is the norm in the burst
limit except for very early times ($\lsim 10^7$ years) and small
$\Mbh$ ($\Mbh\simeq 10^6 \Msun$).

@@

  \begin{enumerate}
  %Younger population=>higher accretion rate; generally higher eta and
  %lower vw except for very young stellar populations.
  \item We find that $\eddr \propto \Mbh^{\alpha}$, where the value of
    $\alpha$ depends on the slope of the stellar density profile
    ($\alpha\simeq0.5,1$ for cusp and core galaxies
    respectively). {\bf AG-these explanations are prehistoric and need
      to be modified. Actually this entire summary needs to be
      revamped, but this can wait until we are closer to completion.}
    One possible solution that the wind mass loss rate (i.e. $\eta$)
    increases with $\Mbh$, as would be expected if low mass SMBH
    preferentially hosted younger stellar populations. Another
    possible explanation is that the effective heating rate of the CNM
    $v_{w}$ decreases with decreasing SMBH mass. This could again be
    due to an older stellar population, since $v_{w} \propto
    \eta^{−1/2}$ for heating sources such as SN Ia and millisecond
    pulsars.
  \item We compare the $\Mdot$ for each of our solution to the one
    calculated from the Bondi formula. We find that they generally to
    a factor of a few. In our model the accretion rate is determined
    from the gas properties at the stagnation radius, $\rs$, which
    agrees with the Bondi radius, $\rb$, within a factor of a
    few. Thus, use of the Bondi formula is a good approximation.
  \item We compare our solutions for density and temperature to those
    inferred from Chandra x-ray observations by \citet{AllenDunn+:2006a}
    and \citet{RussellMcNamara+:2013a} for a handful of galaxies in
    which our sample overlap with theirs. In general, we cannot match
    the observed gas density slope on large scales. 
  \end{enumerate}
